---
title: "Enron EDA"
author: "Giacomo Sarchioni"
date: "23 October 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# Introduction
This document contains the **Exploratory Data Analysis** of the Enron dataset.

```{r}
setwd('/Users/gsarchioni/Desktop/Study/Data Analyst Nanodegree/Lesson6/ud120-projects/submission/eda')
```


```{r}
library(ggplot2)
library(gridExtra)
```

```{r}
df = read.csv('enron.csv')
df$poi = as.factor(df$poi)
```

# Exploration
### Univariate analysis
```{r}
plot_hist <- function(variable) {
  
  ggplot(data = df,
         aes_string(variable)) +
    geom_histogram() +
    theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), 'cm'),
          axis.title = element_text(size = 20))
}
```

```{r}
histograms <- list()
variables = names(df)[-c(1,2,18,22)]

for(i in 1:length(variables)) {
  chart <- plot_hist(variables[i])
  histograms[[i]] <- chart
}

```
It is quite evident how most of the distributions are positively skewed. This makes sense since the variables represent numerical concepts (e.g. salary, bonuses, but even e-mails sent and received) which typically follow such distributions. Most of the observations have medium-low values, while some few cases are far in the right tail.
```{r fig.width=20, fig.height=16}
do.call(grid.arrange,
        c(histograms, ncol = 4))
```

Given the shapes of the distributions, it is worhted to have a look at how transformed distributions (e.g. sqrt and log10) look like. I might do some rescaling at this stage instead of waiting to do it later on during the fitting of features for machine learning algorithms.  
Let me start with **log10** transformations. It is evident how distributions now look more symmetrical.
```{r fig.width=20, fig.height=16}
logs <- list()
variables = names(df)[-c(1,2,18,22)]
for(i in 1:length(variables)) {
  chart <- plot_hist(variables[i])
  chart = chart +
    scale_x_continuous(trans = 'log10')
  logs[[i]] <- chart
}

do.call(grid.arrange,
        c(logs, ncol = 4))
```

I now do the same with **sqrt** transformations. This transformation seems to work very well with the variable salary.
```{r fig.width=20, fig.height=16}
sqrts <- list()
variables = names(df)[-c(1,2,18,22)]
for(i in 1:length(variables)) {
  chart <- plot_hist(variables[i])
  chart = chart +
    scale_x_continuous(trans = 'sqrt')
  sqrts[[i]] <- chart
}

do.call(grid.arrange,
        c(sqrts, ncol = 4))
```

### Bivariate analyis
In this section I will plot some features against each other and comment on some possible relationshisp. This should serve as a kind of "inspirational" exercise in order to inform my decision thinking with regards to the creation of new variables.
```{r}
plot_bivariate <- function(df, indip, depen, xtrans = NULL, ytrans = NULL) {
  
  chart <- ggplot(data = df,
                  aes_string(x = indip, y = depen)) +
    geom_point(aes(color = poi)) +
    scale_color_manual(values = c('blue', 'red'))
  
  if(!is.null(xtrans) & !is.null(ytrans)) {
    chart <- chart +
      scale_x_continuous(trans = xtrans) +
      scale_y_continuous(trans = ytrans)
  } else if(!is.null(xtrans) & is.null(ytrans)) {
    chart <- chart +
      scale_x_continuous(trans = xtrans)
  } else if(is.null(xtrans) & !is.null(ytrans)) {
    chart <- chart +
      scale_y_continuous(trans = ytrans)
  }
  
  return(chart)
  
}
```

#### Salary vs Bonus
The chart below shows the relationship between salary (sqrt-transformed) and bonus (log10-transformed). It is evident how these two variables tell a story. Except for a single observation, all POIs have a bonus above (approximately) 550k USD.
```{r}
chart <- plot_bivariate(df, 'salary', 'bonus', xtrans = 'sqrt', ytrans = 'log10')
chart +
  labs(x = 'salary (sqrt)',
       y = 'bonus (log10)') +
  geom_hline(yintercept = 0.55 * 1e+06,
             linetype = 2,
             size = 0.25) +
  annotate('text',
           x = 900000,
           y = 0.7 * 1e+06,
          label = '0.55M USD')
```

#### Salary vs total stock value
The salary (sqrt-transformed) vs totak stock value (log10-transformed) chart does not seem to bring any significant evidence. 

```{r}
chart <- plot_bivariate(df, 'salary', 'total_stock_value', xtrans = 'sqrt', ytrans = 'log10')
chart + 
  labs(x = 'salary (sqrt)',
       y = 'total stock value (log10)') 
```

#### Messages sent vs received
An interesting chart here. First of all, there seems to be a positive relationship between number of messages sent and number of messages received.  
It also seems that, with the exception of one value, all POIs have received at least 1,000 messags (approximately)
```{r}
chart <- plot_bivariate(df, 'from_messages', 'to_messages', 'log10', 'log10')
chart +
  labs(x = 'sent messages (log10)',
       y = 'received messages (log10)') +
  geom_smooth(colour = 'grey',
              fill = 'light grey',
              alpha = 0.3) +
  geom_hline(yintercept = 1000,
             linetype = 2,
             size = 0.25) +
  theme(panel.background = element_blank(),
        panel.grid.major = element_line(colour = 'grey',
                                        size = 0.1))
```

## New variables
I am now going to explore possible new variables. In particular, I am calculating ratios of existing variables. This will allow to capture me proportios rather than absolute values.

### Financial variables
#### Bonus ratio
This variable expresses the bonus as a proportion of salary.
```{r}
df$bonus_ratio <- with(df,
                       ifelse(!is.na(bonus) & !is.na(salary),
                              bonus/salary,
                              NaN))
```

#### Exercised stock ratio
This variable expresses the value of exercised stock as a proportion of total stock.
```{r}
df$exercised_ratio <- with(df,
                           ifelse(!is.na(exercised_stock_options) &
                                    !is.na(total_stock_value),
                                  exercised_stock_options / total_stock_value,
                                  NaN))
```

#### Payments ratio
The payments_ratio represents the amount of payments as a proportion of salary.  
Spotted another outlier! There is a person with a huge expenses ratio. This person has expenses which are almost 2,000 times his salary.


```{r}
df$payments_ratio <- with(df,
                           ifelse(!is.na(total_payments) &
                                    !is.na(salary),
                                  total_payments / salary,
                                  NaN))
```

```{r}
outlier_payment <- subset(df, payments_ratio > 1500)$name

plot_bivariate(df, 'payments_ratio', 'expenses') +
  annotate('text',
           x = 1600,
           y = 75000,
           label = outlier_payment,
           size = 3)
```

#### Expenses ratio
This variable represents the amount of expenses as a proportion of salary.  
Again, there seems to be an outliere here too, which is the same James Bannantine as before.


```{r}
df$expenses_ratio <- with(df,
                           ifelse(!is.na(expenses) &
                                    !is.na(salary),
                                  expenses/ salary,
                                  NaN))
```

```{r}
outlier_expense <- subset(df, expenses_ratio > 100)$name
```

```{r}
plot_bivariate(df, 'expenses_ratio', 'payments_ratio') +
  annotate('text',
           x = 100,
           y = 1800,
           label = outlier_expense,
           size = 3)
```

#### Wealth variable
I am also adding a simple feature which represents the total wealth of a person as a result of woring/having worked for Enron. This variable is the sum of salary, bonus and total stock value.  
Note that in the final code I will have to transform NaN values into zero, otherwise wealth (as sum of three variables) might not be calculated correctly.


```{r}
# Create temporary salary, bonus and total_stock value
# where nan values are replaced by zero.

df$salary_replaced <- df$salary
df$bonus_replaced <- df$bonus
df$stock_replaced <- df$total_stock_value

# Replace with zero
df$salary_replaced[is.na(df$salary_replaced)] <- 0
df$bonus_replaced[is.na(df$bonus_replaced)] <- 0
df$stock_replaced[is.na(df$stock_replaced)] <- 0

# Add wealth
df$wealth <- with(df, salary_replaced + bonus_replaced + stock_replaced)

# Remove temporary column
drop <- c('salary_replaced', 'bonus_replaced', 'stock_replaced')
df <- df[,!(names(df)) %in% drop]

```


#### Plotting new financial variables
In order to represent these charts, I am going to create a new dataframe where I exclude James Bannantine.
```{r echo = TRUE}
new_df <- subset(df, name != 'BANNANTINE JAMES M')
```

The chart below shows that Kenneth Lay has an extraordinary high payments_ratio (i.e. he received payments of almost 100 times his salary). Since he is a POI, though, I will not remove it from the dataset.
```{r}
name <- subset(new_df, payments_ratio > 50 & !is.na(expenses_ratio))$name
plot_bivariate(new_df, 'payments_ratio', 'expenses_ratio') +
  annotate('text',
           x = 96,
           y = 0.15,
           label = name,
           size = 3)
```



### Messages proportions
For the messagees variable, I am going to create some simple variables which represent the interaction with poi.

#### from_poi_ratio
This variable represents the proportion of e-mail which has been received from a poi out of the total e-mails received.
```{r}
df$from_poi_ratio <- with(df,
                          from_poi_to_this_person / to_messages)
```

#### to_poi_ratio
This variable represents the proportion of e-mail which has been sent to a poi out of the total e-mails sent.
```{r}
df$to_poi_ratio <- with(df,
                        from_this_person_to_poi / from_messages)
```

#### shared_with_poi_ratio
This variable represents the proportion of e-mail which has been received together with one poi at least, ouf of the total e-mails received.
```{r}
df$shared_with_poi_ratio <- with(df,
                                 shared_receipt_with_poi / to_messages)
```

#### Plotting new messages variable
These charts are extremely interesting. If I cut the plot with some horizontal lines, it is evident how poi seem to send at least 23% of their email to other pois. In addition, of all the e-mails they receive, at least 35% of them are shared with poi.  
  
This will be good material to test in the final algorithms.
```{r}
plot_bivariate(df, 'to_poi_ratio', 'shared_with_poi_ratio')
```


